---
title: "Bayesian Assessment of Contemporaneity in Radiocarbon Dates"
subtitle: "A Stan-based Hierarchical Modeling Approach"
author: "Research Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
bibliography: references.bib
---

# Introduction

This document presents a Bayesian hierarchical modeling framework for assessing temporal contemporaneity between archaeological deposits using radiocarbon dates. The analysis uses Stan [@carpenter2017stan] for Bayesian inference and the `rcarbon` package [@crema2021rcarbon] for radiocarbon calibration.

## Objectives

1. Calibrate radiocarbon dates using IntCal20 [@reimer2020intcal20]
2. Fit hierarchical models to assess deposit contemporaneity
3. Compare competing temporal models using LOO cross-validation
4. Quantify overlap probabilities and occupation durations
5. Validate approach using simulation studies

## Theoretical Background

### Radiocarbon Dating and Calibration

Radiocarbon dating measures the decay of ^14^C in organic materials. The measured radiocarbon age must be calibrated to calendar years using a calibration curve that accounts for past atmospheric ^14^C variations [@bronk2009bayesian].

### Contemporaneity Assessment

Statistical contemporaneity does not necessarily imply cultural or behavioral contemporaneity. We define deposits as contemporaneous if their occupation windows overlap in calendar time, accounting for measurement uncertainty.

# Setup

```{r setup}
# Load all required packages and functions
library(rcarbon)
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(dplyr)
library(tidyr)
library(ggplot2)

# Source project functions
source("R/setup.R")
source("R/data_validation.R")
source("R/calibration.R")
source("R/simulation.R")
source("R/stan_utilities.R")
source("R/contemporaneity_metrics.R")
source("R/plotting.R")

# Set random seed for reproducibility
set.seed(42)
```

## Verify Installation

```{r verify-installation}
# Verify Stan installation and package versions
get_package_versions()
verify_stan_installation()
```

# Data

## Data Structure

Our radiocarbon dataset includes the following fields:

- **lab_code**: Laboratory identifier
- **age**: Radiocarbon age (years BP)
- **error**: Measurement error (1σ)
- **deposit**: Deposit identifier
- **material**: Material type (optional)

## Example Dataset

For demonstration, we'll use a simulated dataset with known temporal structure:

```{r generate-example-data}
# Simulate partially overlapping deposits
sim_data <- simulate_partial_overlap(
  n_deposits = 3,
  overlap_percentage = 60,
  deposit_duration = 250,
  earliest_start = 2800,
  n_dates_per_deposit = 10,
  measurement_error = 30,
  seed = 123
)

# Extract radiocarbon data
c14_data <- sim_data$dates[, c("lab_code", "age", "error", "deposit")]

print(sim_data)
```

## Data Validation

```{r validate-data}
# Validate data structure
validate_c14_data(c14_data)

# Print summary
print_c14_summary(c14_data)

# Check for outliers
outliers <- check_outliers(c14_data)
print(outliers[outliers$is_outlier, ])
```

# Calibration

## Calibrate Radiocarbon Dates

```{r calibrate}
# Calibrate using IntCal20
cal_result <- calibrate_dates(
  ages = c14_data$age,
  errors = c14_data$error,
  lab_codes = c14_data$lab_code,
  calCurve = "intcal20",
  verbose = TRUE
)

# Print calibration summary
print(cal_result)
```

## Calibrated Date Distributions

```{r plot-calibrated, fig.width=10, fig.height=8}
# Plot calibrated distributions by deposit
plot_calibrated_dates(cal_result, c14_data)
```

## Calibration Summary Table

```{r calibration-table}
# Create formatted summary table
cal_summary <- cal_result$summary
cal_summary$age <- c14_data$age
cal_summary$error <- c14_data$error
cal_summary$deposit <- c14_data$deposit

knitr::kable(
  cal_summary[, c("lab_code", "deposit", "age", "error",
                  "median_cal_bp", "hdi95_lower", "hdi95_upper")],
  digits = 0,
  col.names = c("Lab Code", "Deposit", "14C Age", "Error",
                "Median (cal BP)", "95% Lower", "95% Upper"),
  caption = "Calibrated Radiocarbon Dates"
)
```

# Bayesian Models

## Model Descriptions

We fit three competing models:

1. **Contemporaneous Model**: All deposits share one occupation window
2. **Sequential Model**: Deposits have separate, non-overlapping windows
3. **Partial Overlap Model**: Deposits have independent windows (most flexible)

## Prepare Stan Data

```{r prepare-stan-data}
# Prepare data for Stan models
stan_data <- prepare_stan_data(c14_data, cal_result)

# Print data structure
str(stan_data)
```

## Compile Stan Models

```{r compile-models}
# Compile all three models
model_contemp <- compile_stan_model("stan/contemporaneous_model.stan")
model_seq <- compile_stan_model("stan/sequential_model.stan")
model_partial <- compile_stan_model("stan/partial_overlap_model.stan")
```

## Fit Models

```{r fit-models, results='hide'}
# Generate initial values
init_contemp <- generate_inits(stan_data, model_type = "contemporaneous")
init_partial <- generate_inits(stan_data, model_type = "partial")

# Fit contemporaneous model
fit_contemp <- fit_stan_model(
  model_contemp,
  stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 42,
  init = init_contemp
)

# Fit sequential model
fit_seq <- fit_stan_model(
  model_seq,
  stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 42,
  init = init_partial
)

# Fit partial overlap model
fit_partial <- fit_stan_model(
  model_partial,
  stan_data,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 42,
  init = init_partial
)
```

## Convergence Diagnostics

```{r diagnostics-contemp}
cat("=== Contemporaneous Model ===\n")
check_stan_diagnostics(fit_contemp)
```

```{r diagnostics-seq}
cat("=== Sequential Model ===\n")
check_stan_diagnostics(fit_seq)
```

```{r diagnostics-partial}
cat("=== Partial Overlap Model ===\n")
check_stan_diagnostics(fit_partial)
```

## Convergence Plots

```{r convergence-plots, fig.width=10, fig.height=6}
# Trace plots for contemporaneous model
conv_plots <- plot_convergence_diagnostics(fit_contemp)
print(conv_plots$trace)
print(conv_plots$rank)
```

# Model Comparison

## LOO Cross-Validation

```{r model-comparison}
# Compare models using LOO
comparison <- compare_models_loo(
  contemporaneous = fit_contemp,
  sequential = fit_seq,
  partial = fit_partial,
  cores = 2
)

# Print comparison
print(comparison$comparison)
print(comparison$weights)
```

## Model Weights Visualization

```{r plot-comparison, fig.width=8, fig.height=5}
plot_model_comparison(comparison)
```

## Interpretation

```{r interpret-comparison}
# Determine best model
best_model <- names(comparison$weights)[which.max(comparison$weights)]
best_weight <- max(comparison$weights)

cat(paste0(
  "The ", best_model, " model has the highest LOO weight (",
  round(best_weight, 3), "), indicating it provides the best ",
  "balance between fit and complexity.\n"
))

if (best_weight > 0.9) {
  cat("Strong evidence for this model structure.\n")
} else if (best_weight > 0.7) {
  cat("Moderate evidence for this model structure.\n")
} else {
  cat("Weak evidence; consider model averaging.\n")
}
```

# Results

## Occupation Boundaries

```{r boundaries, fig.width=10, fig.height=8}
# Using the best-fitting model (partial overlap)
n_deposits <- stan_data$n_deposits

# Plot occupation boundaries
plot_occupation_boundaries(fit_partial, n_deposits)
```

```{r temporal-overlap, fig.width=10, fig.height=6}
# Plot temporal overlap with uncertainty
plot_temporal_overlap(fit_partial, n_deposits)
```

## Occupation Span Estimates

```{r occupation-spans}
# Calculate occupation spans
spans <- calculate_occupation_spans(fit_partial, n_deposits)

knitr::kable(
  spans[, c("deposit", "duration_mean", "duration_sd",
            "start_median", "end_median")],
  digits = 0,
  col.names = c("Deposit", "Duration (mean)", "Duration (SD)",
                "Start (median)", "End (median)"),
  caption = "Occupation Span Estimates (cal BP)"
)
```

## Contemporaneity Metrics

```{r contemporaneity}
# Compute all metrics
metrics <- compute_all_metrics(fit_partial, n_deposits)

# Print summary
print_metrics_summary(metrics)
```

## Overlap Probability Matrix

```{r overlap-matrix, fig.width=8, fig.height=6}
# Visualize pairwise overlap probabilities
plot_overlap_matrix(metrics$overlap_matrix)
```

```{r overlap-table}
# Print as table
knitr::kable(
  metrics$overlap_matrix,
  digits = 3,
  caption = "Pairwise Overlap Probabilities"
)
```

# Model Validation

## Posterior Predictive Checks

```{r ppc, fig.width=10, fig.height=6}
# Check model fit
plot_posterior_predictive(fit_partial, c14_data)
```

## Parameter Estimates vs. Ground Truth

Since we used simulated data with known parameters, we can compare estimates to truth:

```{r compare-to-truth}
# Extract true occupation windows
true_windows <- sim_data$ground_truth$deposit_windows

# Extract estimated windows
estimated_spans <- calculate_occupation_spans(fit_partial, n_deposits)

# Create comparison table
comparison_df <- data.frame(
  deposit = 1:n_deposits,
  true_start = true_windows[, 1],
  est_start = estimated_spans$start_median,
  true_end = true_windows[, 2],
  est_end = estimated_spans$end_median,
  start_error = abs(true_windows[, 1] - estimated_spans$start_median),
  end_error = abs(true_windows[, 2] - estimated_spans$end_median)
)

knitr::kable(
  comparison_df,
  digits = 0,
  caption = "Comparison of True vs. Estimated Boundaries"
)

# Calculate coverage
cat("\nBoundary Coverage Statistics:\n")
for (k in 1:n_deposits) {
  start_covered <- (true_windows[k, 1] >= estimated_spans$start_q2.5[k] &
                    true_windows[k, 1] <= estimated_spans$start_q97.5[k])
  end_covered <- (true_windows[k, 2] >= estimated_spans$end_q2.5[k] &
                  true_windows[k, 2] <= estimated_spans$end_q97.5[k])

  cat(sprintf("Deposit %d: Start %s, End %s\n",
              k,
              ifelse(start_covered, "COVERED", "NOT COVERED"),
              ifelse(end_covered, "COVERED", "NOT COVERED")))
}
```

# Discussion

## Summary of Findings

1. **Model Selection**: The {best_model} model provides the best fit
2. **Contemporaneity**: Overlap probabilities quantified for all deposit pairs
3. **Validation**: Estimates show good agreement with known ground truth

## Limitations

- Statistical contemporaneity ≠ cultural contemporaneity
- Assumes uniform sampling within occupation windows
- Sensitive to measurement error and sample size
- Does not account for reservoir effects or contamination

## Recommendations

**Sample Size Guidelines:**

- Minimum 5-10 dates per deposit for reliable estimates
- More dates needed when deposits are close in time
- Consider power analysis for study design

**Model Selection:**

- Use LOO weights > 0.7 for confident model selection
- Consider model averaging when weights are diffuse
- Sequential model requires strong prior evidence

# References

::: {#refs}
:::

# Session Info

```{r session-info}
sessionInfo()
```

# Appendix: Code Repository

All code, Stan models, and data are available at:
[GitHub Repository Link]

To reproduce this analysis:

```bash
# Clone repository
git clone [repository-url]

# Install dependencies
Rscript -e "source('R/setup.R'); complete_setup()"

# Run analysis
quarto render analysis.qmd
```
